---
title: "Street Address Data for Victoria"
author: "Dhirendra Singh"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
  beamer_presentation: default
  pdf_document:
    highlight: tango
  html_document:
    keep_md: yes
urlcolor: blue
#geometry: "left=1cm,right=1cm,top=2cm,bottom=2cm"
fontsize: 10pt
---

## About 

*Document version: `r as.character(format(Sys.Date(), format="%B %d, %Y"))`*

This document describes the process of creating a readily useable CSV file of all street addresses in the state of Victoria. We use the Vicmap Address dataset for this purpose.

# Data

## VicMap Data

The `Vicmap Address` dataset for the state of Victoria was downloaded from the [Spatial Datamart](http://services.land.vic.gov.au/SpatialDatamart) service in the following format:

--------------------------------------------------------------------------
Product         Area	                Format	         Projection
--------------- --------------------- ---------------- -------------------
Vicmap Address  Whole of State : VIC	ESRI Shape file	 VICGRID on GDA-94
--------------------------------------------------------------------------

Given its size (268M) the archive is not saved in this repository, but can be downloaded from [this link](https://cloudstor.aarnet.edu.au/plus/s/efejsmkNL6XCM77).

# Data Processing

## Extract VicMap Data

Unzip the Vicmap archive. Note that ESRI Shapefile databases (`.dbf`) have a size limit of 2GB, due to which Vicmap supplies the GIS addresses for Victoria cut vertically into two files.

```{r, engine = 'bash', eval = FALSE}
  $ unzip SDM596803.zip
  Archive:  SDM596803.zip
  inflating: Data_Access_Licence.pdf  
  inflating: VICMAP_ADDRESS.xml      
  inflating: ANZVI0803002578.mhtml   
  inflating: ANZVI0803003126.mhtml   
   creating: vicgrid94/
   creating: vicgrid94/shape/
   creating: vicgrid94/shape/whole_of_dataset/
   creating: vicgrid94/shape/whole_of_dataset/vic/
   creating: vicgrid94/shape/whole_of_dataset/vic/VMADD/
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS_column_names.txt  
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS.shp  
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS.shx  
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS.dbf  
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS.prj  
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS.cpg  
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS_1.shp  
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS_1.shx  
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS_1.dbf  
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS_1.prj  
  inflating: vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS_1.cpg  
```

## ESRI to GeoJson

Convert the ESRI shapefiles to compressed GeoJSON files in WSG84 format for easier processing:

```{r, engine = 'bash', eval = FALSE}
ogr2ogr -f "GeoJson" add.json vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS.shp
gzip add.json

ogr2ogr -f "GeoJson" -t_srs "EPSG:4326" add1.json vicgrid94/shape/whole_of_dataset/vic/VMADD/ADDRESS_1.shp
gzip add1.json
```

Note that the process takes a few minites and the interim `.json` files are quite large (up to 8GB), due to the volume of data. We final compressed GeoJson files are more manageable:

```{r, engine = 'bash', eval = FALSE}
$ du add*json.gz
305M	add.json.gz
3.1M	add1.json.gz
```

## GeoJson to CSV

Extract the fields `EZI_ADD`, `MESH_BLOCK`, and `coordinates` from the GeoJson files into a combined compressed CSV:

```{r, engine = 'bash', eval = FALSE}
$ zless add.json.gz |\
    grep -oE "EZI_ADD.*?,|MESH_BLOCK.*?,|coordinates.*?]" |\
    cut -f2 -d':' |\
    sed -e 's/\[//g;s/\]//g' |\
    awk '{$1=$1};1'|\
    paste -s -d ",,\n" - |\
    sed -e 's/,,/,/g' > \
    add.csv
    
$ zless add1.json.gz |\
    grep -oE "EZI_ADD.*?,|MESH_BLOCK.*?,|coordinates.*?]" |\
    cut -f2 -d':' |\
    sed -e 's/\[//g;s/\]//g' |\
    awk '{$1=$1};1'|\
    paste -s -d ",,\n" - |\
    sed -e 's/,,/,/g' > \
    add1.csv
    
$ (echo "EZI_ADD,MESH_BLOCK,X,Y" && cat add1.csv) | gzip > VICMAP-ADDRESSES-2019-VIC.csv.gz
```

## Output Street Address Records

Below are some sample records (rows) from the extracted `VICMAP-ADDRESSES-2019-VIC.csv.gz` file. 

-------------------------------------------------------------------------------------------------
EZI_ADD                                 MESH_BLOCK     X                    Y
--------------------------------------- -------------- -------------------  ---------------------
"123 FOREST STREET BENDIGO 3550"        "20206450000"  144.27175086         -36.755240929999722

"5-9 HELEN STREET MOUNT DANDENONG 3767" "20649530000"  145.358282319999944  -37.834493579999759

"LAE STREET WEST FOOTSCRAY 3012"        "20379840000"  144.866444788999956  -37.809529774999767

"13 DEANS MARSH ROAD LORNE 3232"        "20562580000"  143.97666909         -38.533417849999772

"8A EASTSIDE DRIVE MILDURA 3500"        "20411621000"  142.184774589999961  -34.195369369999646
-------------------------------------------------------------------------------------------------


## Street Addresses to Statistical Areas

We link street addresses to Statistical Areas using the [`Victoria Mesh Blocks ASGS Edition 2016`](https://www.abs.gov.au/AUSSTATS/abs@.nsf/Lookup/1270.0.55.001Main+Features10018July%202016?OpenDocument) dataset from the ABS, downloaded on 16 May 2019 as `../abs/MB_2016_VIC.csv.gz`. 

Here are partial rows showing the mapping to Statistical Areas from Mesh Bocks:

---------------------------------------------------------------------------------------------------------------------------
MB_CODE      SA1_7DIGITCODE SA2_5DIGITCODE SA2_NAME         SA3_CODE SA3_NAME             SA4_CODE  SA4_NAME
_2016        _2016          _2016          _2016            _2016    _2016                _2016     _2016
------------ -------------- -------------- ---------------- -------- -------------------- --------- --------------
20206450000  2101822        21018          Bendigo          20201    Bendigo              202       Bendigo

20649530000  2128221        21282          Mount Dandenong  21105    Yarra Ranges         211       Melbourne -
                                           - Olinda                                                 Outer East
                                           
20562580000  2104902        21049          Lorne -          20303    Surf Coast -         203       Geelong
                                           Anglesea                  Bellarine Peninsula
---------------------------------------------------------------------------------------------------------------------------

## Missing Mesh Blocks?

* Vicmap data in (`VICMAP-ADDRESSES-2019-VIC.csv.gz`) has `3,770,081` street addresses 
belonging to `77,592` Mesh Blocks. 

* In the ABS data (`MB_2016_VIC.csv.gz`) there are `85,014` Mesh Block records.

* However, some Vicmap Mesh Blocks, such as `20379840000` and `20411621000` from before, do not exist in the ABS dataset. 

* According to the [Vicmap Address Product Description](https://www2.delwp.vic.gov.au/maps/spatial-data?a=49746) document the `mesh_block` attribute is `Not currently populated` possibly suggesing that it should not be used (yet).

* One option is to find which mesh block polygon the coordinates of the address record belongs to. We have a fast algorithm for this, called [Address Mapper](https://github.com/agentsoz/synthetic-population/tree/v2#address-mapper) that can be used.


## Residential Street Addresses

The ABS Mesh Block records also contain the `MB_CATEGORY_NAME_2016` field which has some useful classification that could be used to categorise street addesses in those blocks:

```
-----------------------------------------------------
Category            Number of Mesh Blocks in Category
------------------- ---------------------------------
Residential         58559 
Parkland            10551 
Primary Production  4634 
Commercial          4309 
Industrial          2122 
Education           2057 
Other               1112 
Water               810 
Transport           595 
Hospital/Medical    257 
SHIPPING            5 
OFFSHORE            1 
NOUSUALRESIDENCE    1 
MIGRATORY           1 
------------------------------------------------------
```